Ignition of Fuel Beds by Firebrands

### 2022-06-17 manuscript2.tex
Finished adding CFD parts.

      The inlet boundary conditions were determined from experimental measurements where the bulk wind speed was 5.8 \si{\meter\per\second} and the inlet turbulence intensity was 0.14\%. The solid surfaces (i.e., cartridge heater, holding rods, fuel bed, and wind tunnel floor) were considered with a no-slip boundary condition and at fixed temperatures. All temperatures except the heater surfaces and pyrolyzate injection were 300\si{\kelvin}. The cartridge heater was maintained at a fixed uniform temperature corresponding to the temperature estimated to result in 50\% ignition probability from experimental tests. The inlet temperature and mass flux of the pyrolysis gases were determined from the fuel bed thermal model and Cantera calculations.  The outlet and sides of the domain used a zero gradient boundary condition. The local equivalence ratio was calculated after one second of pyrolysis release. While these times are much less than the average time to ignition, insights are still gained into the distribution of pyrolyzates with respect to the observed ignition locations for the heater orientations.

>>>>

      The inlet boundary conditions were determined from experimental measurements where the bulk wind speed was 5.8 \si{\meter\per\second} and the inlet turbulence intensity was 0.14\%. The solid surfaces (i.e., cartridge heater, holding rods, fuel bed, and wind tunnel floor) were considered with a no-slip boundary condition and at fixed temperatures. All temperatures except the heater surfaces and pyrolyzate injection were 300\si{\kelvin}. The cartridge heater was maintained at a fixed uniform temperature corresponding to the temperature estimated to result in 50\% ignition probability from experimental tests. The inlet temperature and mass flux of the pyrolysis gases were determined from the fuel bed thermal model and Cantera calculations.  The outlet and sides of the domain used a zero gradient boundary condition. The local equivalence ratio was calculated after one second of pyrolysis release. While these times are much less than the average time to ignition, insights are still gained into the distribution of pyrolyzates with respect to the observed ignition locations for the heater orientations. For validation the separation angle of the flow was compared to a correlation from literature~\cite{Jiang2020SeparationRegime} and was found to predict the angle within 1\% of the recommended correlation. Further grid refinement did not appreciably reduce the error and resulted in a significant increase in computational time. Thus, the calculations were considered of acceptable accuracy to visualize the trends observed in experiments.

### 2022-06-17 manuscript4.tex
Finished adding CFD parts.

\section{Methodology}
    The probability of flaming ignition for fuel beds consisting of Douglas-fir particles was measured for eight configurations of two firebrands in a wind tunnel. The wind tunnel was operated at either 0.5 \si{\meter\per\second} or 5.8 \si{\meter\per\second} for each test series. The wind speed was measured with a TSI-IFA300 hot wire probe. Figure~\ref{fig:multiHeaterApparatus} shows a representation of the wind tunnel, fuel bed, and the automated lowering device. Table~\ref{tab:multiHeaterConfig} shows the test matrix for each series of tests. For all of the tests the heaters were oriented perpendicular to the flow and the downstream heater was heated. For tests where both heaters were heated the temperatures of both heaters were maintained at the same temperature.

@@

@@

    The thermal model of the fuel bed was conducted using OpenFOAM~\cite{Foundation2020}. Figure~\ref{fig:multiHeaterThermal} shows the computational domain of the fuel bed. The thermal diffusivity of the fuel bed materials was estimated using correlations from literature for the thermal conductivity~\cite{bergman2011fundamentals} and published values of specific heat~\cite{Parker1986}. Equation~\ref{eqn:estk4} shows the correlation for effective thermal conductivity where $\epsilon$, as defined in Equation~\ref{eqn:epsilon4}, is the proportion of the fuel bed that is air. The bulk density of the fuel bed material ($\rho_{bed}$) and the solid wood ($\rho_{solid}$) were obtained from experimental samples. 
        \begin{equation}
            k_{eff} = \frac{1}{2} \left(\frac{1}{\left( 1 - \epsilon \right)/k_{solid} + \epsilon/k_{air}} +  \epsilon k_{air} + \left(1-\epsilon \right) k_{solid} \right)
            \label{eqn:estk4}
        \end{equation}
        \begin{equation}
            \epsilon = 1- \frac{\rho_{solid}}{\rho_{bed}}
            \label{eqn:epsilon4}
        \end{equation}
    The interface between the fuel bed and air and the radial edge of the domain were treated as insulated. The centerline was treated as a symmetric boundary condition. The insulated radial edge and symmetric boundary conditions are consistent with boundary conditions in the actual experiment since the 25\si{\milli\meter} distance ensure no temperature change during 10\si{\second} of heating modeled, and the heater temperature has been observed to be circumferentially consistent with infrared imaging. The insulated interface with the air does not account for radiation and convective heat losses from the surface of the fuel bed as it is heated. However, the magnitude of the heat losses from the fuel bed is estimated to be significantly smaller in magnitude than the heat transfer from the heater. Reactions, mass loss due to pyrolysis, and changes in contact area due to changing fuel bed geometry are not considered in the temperature analysis. While the aforementioned assumptions and limitations reduce the accuracy of the model in comparison to the experimental conditions, the temperature distributions still provide insights into differences in the mass of the fuel bed material that undergoes pyrolysis and provides insight to better interpret the experimental results. 
    \begin{figure}
        \centering
        \resizebox{0.375\textwidth}{!}{%
          \begin{tikzpicture}
            \node at (0,0) (origin) {};
            %\draw (origin) rectangle (topright);
            \fill[pattern=north west lines, pattern color= brown, draw=black] (3.18mm, 0) -- (25mm, 0) arc (0:-90:25mm) -- (0, -3.18mm) arc (270:0:3.18mm ) -- cycle ;

@@

    The velocity of pyrolyzates entering the fluid domain from the fuel bed was determined in a two-step process. First, the mass of material in the fuel bed undergoing pyrolysis was estimated. All material over a temperature of 220\si{\celsius} was considered to undergo pyrolysis. This temperature was chosen as 220\si{\celsius} corresponds to the minimum temperature for the pyrolysis of hemicellulose and is considered the onset of pyrolysis~\cite{Yang2007a}. The amount of mass estimated to depart the fuel bed was determined to be the proportion of fuel bed material that converted to gas-phase products when reacted at the average temperature of the material above the pyrolysis threshold from the thermal model. Reactions were conducted using the BioPOx mechanism~\cite{Dhahak2019}, and mass was assumed to depart the fuel bed if it existed in the gas phase. Species were considered to be gas-phase if they were present in the Bio1412 mechanism~\cite{Ranzi2001, Ranzi2008} since the Bio1412 mechanism considers the gas-phase reaction of pyrolysis products. The reaction calculations were conducted in Cantera~\cite{Goodwin2020}. The Cantera calculations were 0D which necessitated using the average temperatures of the fuel bed for the pyrolysis calculations rather than spatially resolved temperatures. Mass flux of the gases exiting the fuel bed was determined by averaging the total mass of gaseous pyrolysis products produced over the 10\si{\second} interval. The exit velocity was then determined as the mass flux through the distance between the heater edge and the pyrolysis threshold, as shown by the dotted line in Figure~\ref{fig:multiHeaterThermal}. The calculated velocity and injection area was then used as an input for calculations conducted in the flow field.

    A 2D RANS model of the flow above the fuel bed and around the heaters was implemented in OpenFOAM. The area of the computational domain with respect to the experimental apparatus is shown as the dashed line in Figure~\ref{fig:multiHeaterApparatus}. The 2D slice of the domain modeled was considered to pass through the center of the heaters and corresponds to the location of the highest temperature of the heater, which also typically coincides with the ignition location. Turbulence parameters were determined from measurements taken of the experimental apparatus with a TSI IFA-300 hot wire anemometer. The k-$\epsilon$ turbulence model was used with values of 1.4$\cdot$10\textsuperscript{-4}\si{\square\meter\per\square\second} and 9.5$\cdot$10\textsuperscript{-6}\si{\square\meter\per\cubic\second} for k and $\epsilon$ respectively. Calculations were conducted for 10\si{\second} and were then averaged to observe differences in temperature and velocity characteristics between the different configurations.

>>>>

\section{Methodology}
    \subsection{Experimental}
    The probability of flaming ignition for fuel beds consisting of Douglas-fir particles was measured for eight configurations of two firebrands in a wind tunnel. The wind tunnel was operated at either 0.5 \si{\meter\per\second} or 5.8 \si{\meter\per\second} for each test series. The wind speed was measured with a TSI-IFA300 hot wire probe. Figure~\ref{fig:multiHeaterApparatus} shows a representation of the wind tunnel, fuel bed, and the automated lowering device. Table~\ref{tab:multiHeaterConfig} shows the test matrix for each series of tests. For all of the tests the heaters were oriented perpendicular to the flow and the downstream heater was heated. For tests where both heaters were heated the temperatures of both heaters were maintained at the same temperature.

@@

    \subsection{Computational}

@@

    The thermal model of the fuel bed was conducted using OpenFOAM~\cite{Foundation2020}. The calculations solved the heat diffusion equation (Equation~\ref{eqn:heatDiffusion4} to observe the spatial and temporal evolution of the fuel bed temperatures.
        \begin{equation}
            \frac{\partial \left(\rho h \right)}{\partial t} = \frac{\partial}{\partial x_{j}} \left( \alpha \frac{\partial h}{\partial x_{j}} \right)
            \label{eqn:heatDiffusion4}
        \end{equation}

    Figure~\ref{fig:multiHeaterThermal} shows the computational domain of the fuel bed. The thermal diffusivity of the fuel bed materials was estimated using correlations from literature for the thermal conductivity~\cite{bergman2011fundamentals} and published values of specific heat~\cite{Parker1986}. Equation~\ref{eqn:estk4} shows the correlation for effective thermal conductivity where $\epsilon$, as defined in Equation~\ref{eqn:epsilon4}, is the proportion of the fuel bed that is air. The bulk density of the fuel bed material ($\rho_{bed}$) and the solid wood ($\rho_{solid}$) were obtained from experimental samples. 
        \begin{equation}
            k_{eff} = \frac{1}{2} \left(\frac{1}{\left( 1 - \epsilon \right)/k_{solid} + \epsilon/k_{air}} +  \epsilon k_{air} + \left(1-\epsilon \right) k_{solid} \right)
            \label{eqn:estk4}
        \end{equation}
        \begin{equation}
            \epsilon = 1- \frac{\rho_{solid}}{\rho_{bed}}
            \label{eqn:epsilon4}
        \end{equation}
    Average cell sizes for the domain ranged between 0.02\si{\milli\meter} near the heater to 0.2\si{\milli\meter} near the insulated boundary. The mesh consisted of 500,000 cells. The interface between the fuel bed and air and the radial edge of the domain were treated as insulated. The centerline was treated as a symmetric boundary condition. The insulated radial edge and symmetric boundary conditions are consistent with boundary conditions in the actual experiment since the 25\si{\milli\meter} distance ensure no temperature change during 10\si{\second} of heating modeled, and the heater temperature has been observed to be circumferentially consistent with infrared imaging. The insulated interface with the air does not account for radiation and convective heat losses from the surface of the fuel bed as it is heated. However, the magnitude of the heat losses from the fuel bed is estimated to be significantly smaller in magnitude than the heat transfer from the heater. Reactions, mass loss due to pyrolysis, and changes in contact area due to changing fuel bed geometry are not considered in the temperature analysis. While the aforementioned assumptions and limitations reduce the accuracy of the model in comparison to the experimental conditions, the temperature distributions still provide insights into differences in the mass of the fuel bed material that undergoes pyrolysis and provides insight to better interpret the experimental results. 
    \begin{figure}
        \centering
        \resizebox{0.375\textwidth}{!}{%
          \begin{tikzpicture}
            \node at (0,0) (origin) {};
            %\draw (origin) rectangle (topright);
            \fill[pattern=north west lines, pattern color= brown, draw=black] (3.18mm, 0) -- (25mm, 0) arc (0:-90:25mm) -- (0, -3.18mm) arc (270:0:3.18mm ) -- cycle ;

@@

    The velocity of pyrolyzates entering the fluid domain from the fuel bed was determined in a two-step process. First, the mass of material in the fuel bed undergoing pyrolysis was estimated. All material over a temperature of 220\si{\celsius} was considered to undergo pyrolysis. This temperature was chosen as 220\si{\celsius} corresponds to the minimum temperature for the pyrolysis of hemicellulose and is considered the onset of pyrolysis~\cite{Yang2007a}. The amount of mass estimated to depart the fuel bed was determined to be the proportion of fuel bed material that converted to gas-phase products when reacted at the average temperature of the material above the pyrolysis threshold from the thermal model. Reactions were conducted using the BioPOx mechanism~\cite{Dhahak2019}, and mass was assumed to depart the fuel bed if it existed in the gas phase. Species were considered to be gas-phase if they were present in the Bio1412 mechanism~\cite{Ranzi2001, Ranzi2008} since the Bio1412 mechanism considers the gas-phase reaction of pyrolysis products. The reaction calculations were conducted in Cantera~\cite{Goodwin2020}. The Cantera calculations were 0D which necessitated using the average temperatures of the fuel bed for the pyrolysis calculations rather than spatially resolved temperatures. The Cantera calculations considered the fuel bed material as an insulated fixed mass reactor. The following equations were solved to obtain the temperature and species concentrations of the pyrolyis products. 
        \begin{equation}
            \pder[m]{t} = 0
        \end{equation}

        \begin{equation}
            m \pder[\left( m Y_k \right)]{t} = \dot{m}_{k, gen} = V \dot{\omega}_k MW_k
        \end{equation}
        \begin{equation}
            mc_{v} \pder[T]{t} = \dot{Q} - \sum_{k} \dot{m}_{k,gen}u_k
        \end{equation}
    Where $Y_k$, $\dot{m}_{k}$, and  $\dot{\omega}_k$ are the mass fraction, mass generation rate, and generation rate of each species $k$ in the mechanism used. $V$ is the volume of the fuel bed material in the reactor. The species generation rates $ \dot{\omega}_k$ are determined from the reactions included in the mechanism. The mechanism used for pyrolysis contains four different types of chemical reactions. The equations used to calculate the reaction rates for an elementary reaction is:
        \begin{align}
            k_f &= AT^{b}e^{-E_a/RT}\\
            R_{f} &= [A][B]k_f 
        \end{align}
    Where $R_f$ is the forward reaction rate, $k_f$ the reaction rate constant, $E_a$ the activation energy, $b$ is the temperature exponent and $R$ is the gas constant. Similarly, the reaction rates for three body reactions are calculated as:
        \begin{align}
            R_f &= [A][B][M]k_f \\
            [M] &= \sum_{k}\epsilon_k C_k
        \end{align}
    Where $\epsilon_k$ and $C_k$ are the collision efficiency and concentration of each species, respectively. Fallof reaction rates are calculated using the reduced pressure $P$ defined as:
        \begin{equation}
            P = \frac{k_0[M]}{k_\infty}
        \end{equation}
    The reaction reaction rate is then calculated as:
        \begin{equation}
            k_f (T, P) = k_{\infty} \left( \frac{P}{1+P}\right) F(T,P)
        \end{equation}
      Pressure dependent P-Log reaction rates are calculated for intermediate values using a logarithmic interpolation between two reaction rates $k_1$ and $k_2$ at pressures $P_1$ and $P_2$ as:
        \begin{equation}
            \log k_f(T,P) = \log k_1(T) + (\log k_2(T) - \log k_1(T))\frac{\log P - \log P_1}{\log P_2 - \log P_1}
        \end{equation}
    The overall generation rate of each species ($\dot{\omega}_k$) is then defined as the sum of the individual reaction rates $R_f$ for all of the reactions where the species is present. 
        \begin{equation}
            \dot{\omega}_k = \sum_i^N \nu_i R_{f, i}
        \end{equation}
    Where $\nu_i$ number of each species produced for each reaction.

    The mass flux of the gases exiting the fuel bed was determined by averaging the total mass of gaseous pyrolysis products produced over the 10\si{\second} interval. The exit velocity was then determined as the mass flux through the distance between the heater edge and the pyrolysis threshold, as shown by the dotted line in Figure~\ref{fig:multiHeaterThermal}. The calculated velocity and injection area was then used as an input for calculations conducted in the flow field.

    A 2D RANS model of the flow above the fuel bed and around the heaters was implemented in OpenFOAM. The area of the computational domain with respect to the experimental apparatus is shown as the dashed line in Figure~\ref{fig:multiHeaterApparatus}. The 2D slice of the domain modeled was considered to pass through the center of the heaters and corresponds to the location of the highest temperature of the heater, which also typically coincides with the ignition location. The  mass, momentum, species, and energy equations for the calculations are shown below~\cite{Versteeg2007AnMethod}.
            \begin{equation}
                \pder[\bar{\rho}]{t} + \pder{x_i} (\bar{\rho} \tilde{u_i}) = 0
            \end{equation}

            \begin{equation}
                \pder[(\bar{\rho} \tilde{u_i})]{x_t} + \pder{x_i} (\bar{\rho} \tilde{u_i} \tilde{u_j}) + \pder[\bar{p}]{x_j} = \pder{x_i} (\bar{\tau}_{ij} -\bar{\rho}\widetilde{u''_i u''_j})
            \end{equation}

            \begin{equation}
                \pder[\bar{(\rho} \tilde{Y_k})]{t} + \pder{x_i} (\bar{\rho} \tilde{u_i} \tilde{Y_k}) = - \pder{x_i} \left(\overline{V_{k,i}Y_{k}} + \bar{\rho} \widetilde{u''_{i} Y''_{k}}\right) +  \dot{\omega}_k
            \end{equation}

            \begin{multline}
                \pder[\bar{\rho}\tilde{h_s}]{t} + \pder{x_i} (\bar{\rho} \tilde{u_i} \tilde{h_s}) = \\ \bar{\dot{\omega_T}} + \frac{\bar{Dp}}{Dt} + \pder{x_i} \left( \overline{\lambda \pder[T]{x_i}} - \overline{\rho u''_i h''_s} \right) + \overline{\tau_{ij} \pder[u_i]{x_j}} - \pder{x_i} \left( \overline{\rho \sum_{k=1}^n V_{k,i} Y_k h_{s,k}} \right)
            \end{multline}
        where
            \begin{equation}
                \frac{\bar{Dp}}{Dt} = \pder[\bar{p}]{t} + \tilde{u_i} \pder[\bar{p}]{x_i}+ \overline{u''_i \pder[p]{x_i}}    
            \end{equation}
     The domain used for all calculations performed was 125\si{\milli\meter} long and 75\si{\milli\meter} high. For the cases where the heaters were spaced one diameter apart a total of 13206 cells with characteristic lengths from 0.05\si{\milli\meter} in the region surrounding the heaters to 1.5\si{\milli\meter} in the free stream above the heaters. For the cases where the heaters were spaced five diameters apart a total of 21948 cell with characteristic lengths from 0.06\si{\milli\meter} in the region surrounding the heaters to 1.6\si{\milli\meter} in the free stream above the heaters. For both cases the fine mesh spacing extended between the heaters accounting for the increased cell count for the five diameter spaced case. Further refinement of the mesh resulted in a shift in separation angle for the upstream heater of 0.2\%. Thus, the calculations were considered of acceptable accuracy to visualize the trends observed in experiments.

    The inlet boundary conditions were determined from experimental measurements where the bulk wind speeds were 5.8 \si{\meter\per\second}, 3.5 \si{\meter\per\second}, 0.5 \si{\meter\per\second} with corresponding inlet turbulence intensities of 0.14\%, 0.2\% and 1.6\% respectively. Turbulence parameters were determined from measurements taken of the experimental apparatus with a TSI IFA-300 hot wire anemometer. The k-$\epsilon$ turbulence model was used with values of 1.4$\cdot$10\textsuperscript{-4}\si{\square\meter\per\square\second} and 9.5$\cdot$10\textsuperscript{-6}\si{\square\meter\per\cubic\second} for k and $\epsilon$ respectively. Calculations were conducted for 10\si{\second} and were then averaged to observe differences in temperature and velocity characteristics between the different configurations. The solid surfaces (i.e., cartridge heater, fuel bed, and wind tunnel floor) were considered with a no-slip boundary condition at fixed temperatures. All temperatures except the heater surfaces were 300\si{\kelvin}. The inlet temperature and inlet velocity of the pyrolysis gases were determined from the fuel bed thermal model and Cantera calculations. The cartridge heater was maintained at a fixed uniform temperature corresponding to the temperature estimated to result in 50\% ignition probability from experimental tests. The outlet and tap of the domain used a zero gradient boundary condition.
